Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2019-02-20 paper/manuscript.tex
Assigning sections

\author[1*]{John D. Chodera}
\author[2*]{Antonia S. J. S. Mey}
\author[2*]{Julien Michel}
\author[3*]{David L. Mobley}
\author[4*]{Conor Parks}
\author[5*]{Julia E. Rice}
\author[6*]{Michael Shirts}
\author[7]{Bryce K. Allen}
\author[1]{Levi N. Naden}
\author[1,9]{Andrea Rizzi}
\affil[1]{Computational and Systems Biology Program, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[2]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[3]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
\affil[4]{I don't know my affiliation}
\affil[5]{I don't know my affiliation}
\affil[6]{University of Colorado Boulder, Boulder, CO, USA}
\affil[7]{Silicon Therapeutics, Boston, MA, USA}
\affil[8]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}

@@

%%%%%%%%%%%%%%%%%%%%
%              Introduction                  %
%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
Alchemical free energy calculations have become a mature technology for computing various properties related to the transfer of chemical species from one environment to another.
The domain of applicability for these calculations now involves such varied applications as the computation of protein-ligand binding free energies~\cite{binding-free-energies}, ligand selectivities~\cite{selectivity}, partition and distribution coefficients between different liquid phases (such as octanol-water partition coefficients~\cite{octanol-water-partition}), loss of affinity due to resistance mutations~\cite{resistance-mutations}, and changes in protein thermostability due to engineered mutations~\cite{thermostability}.

The defining characteristic of alchemical free energy calculations is the use of a series of modified potential functions $U(x; \lambda)$ in which an alchemical parameter $\lambda$ modulates interactions in a manner that cannot occur in real chemical systems.
One or more simulations are used to collect data from a multitude of alchemical states to compute a free energy difference between a chemical state ($\lambda_0$) and another chemical or alchemical reference state ($\lambda_1$),
\begin{eqnarray}
\Delta f &\equiv& f(\lambda_1) - f(\lambda_0) = - \ln \frac{Z(\lambda_1)}{Z(\lambda_0)}
\end{eqnarray}
where the dimensionless free energy $f(\lambda) \equiv \beta F(\lambda)$ is given in terms of partition functions $Z(\lambda)$,
\begin{eqnarray}
Z(\lambda) &=& \int dx \, e^{-u(x; \lambda)} .
\end{eqnarray}
Here, the inverse temperature $\beta \equiv (k_B T)^{-1}$ where $k_B$ is the Boltzmann constant, and the \emph{reduced potential} $u(x; \lambda)$~\cite{reduced-potential} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
\begin{eqnarray}
u(x;\lambda) &\equiv& \beta \left[ U(x;\lambda) + p \, V(x) + \sum_{i=1}^N \mu_i N_i(x) + \cdots \right]
\end{eqnarray}
where the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
The physical transformation of interest may require several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity.

@@

Before proceeding with this document, you are assumed to have a basic familiarity with the principles of molecular simulations, molecular dynamics simulations, statistical mechanics and thermodynamics, and so on.
This document focuses on aspects specific to alchemical free energy calculations, which remain an advanced topic within the molecular simulations area.
Thus, if you are a beginner in the area of molecular simulations, not all background you need will be covered within the scope of this document.
For example, we assume that you are able to set up and conduct a successful equilibrium molecular dynamics simulation of a particular thermodynamic state of a system interest as a prerequisite for this document.

@@

\section{Step 0 -- What can be expected from alchemical simulations?}
\label{sec:step0}
\begin{itemize}
\item What level of accuracy you can expect?
\item What timescales and how many transformations can you address given available computational resources?
\item Can you even hope to tackle the problem you are attempting?
\end{itemize}

\section{Step 1 -- Simulation prerequesites}
\label{sec:step1}
\begin{enumerate}
\item Generate geometry of initial state: Reference biomolecular simulation preparation best practices
\item Relative: Generate geometry of final state (Mey)
\begin{itemize}
\item e.g. ligand ideally should not clash with receptor, etc.; satisfy constraints that might be imposed by protocol such as overlapping atoms, etc.
\item reference biomolecular preparation setup practices for placing ligand into binding site, but elaborate on constraints that must be considered for relative free energy calculations
\end{itemize}
\end{enumerate}

@@

Many protocol issues for alchemical calculations are common, but some are different between absolute and relative calculations, so before treating the common elements we treat the protocol differences.

\subsubsection{Relative free energy calculations must select a topology and produce an atom mapping}

@@

\section{Step 3 -- Overview of available analysis techniques}
\label{sec:step3}
\todo[inline, color={red!40}]{JDC, BA: uncertainty estimation section}
\begin{enumerate}
\item Detecting boundary between equilibrated and production regions (Chodera: \url{http://dx.doi.org/10.1021/acs.jctc.5b00784})
\item Decorrelating samples for analysis
\begin{itemize}
\item Subsample different lambdas based on correlation times
\item Ensure all simulations at least 50x correlation time
\end{itemize}
\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
\begin{itemize}
\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
\item Poor replica mixing (for replica-exchange)
\item Correlation time as a function of lambda as it would be expected to be a smooth
\item Dependence on initial conformation
\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
\item More “usual suspects”
\end{itemize}
\item Estimators for free energies
\begin{itemize}
\item MBAR recommended if all energy differences are available
\item BAR just as good for highly optimized lambda values
\item TI should be roughly concordant, but quadrature error hard to quantify
\item Other variants useful in special circumstances (e.g. Z. Tan stochastic version)
\end{itemize}
\item Computing and reporting uncertainties on free energies
correlated bootstrap v. timeseries analysis
\begin{itemize}
\item Quantifying standard error in dG estimate
\item Varience in dG estimate using multiple methods (TI, DEXP, IEXP, BAR, MBAR, GDEL, etc.)
\item Agreement in dG estimate when repeating calculations with different parameters (random seeds, initial configurations, forcefield, etc.)
\item Calculating differences in free energy change as a function of time
\begin{itemize}
	\item Starting from beginning or end of simulation
	\item Significance of differences in midpoint estimate (High middle error: high uncertainty)
\end{itemize}
\item Ensemble method to combine uncertainties into interpretable weighted metric
\begin{itemize}
	\item Simple normalized metric to determine confidence in calculation
	\item Easily interpreted by chemist/biologist when prioritizing new chemistries
\end{itemize}
\end{itemize}
\item Other considerations for many transformations
Cycle closure error

@@

%%%%%%%%%%%%%%%%%%%%
%              Terminology                  %
%%%%%%%%%%%%%%%%%%%%
\section{Terminology and abbreviations}
\label{sec:tem-abbrev}
\begin{itemize}
\item Feature Box covering major technical terms and abbreviations
\item Examples:
\begin{itemize}
\item EXP, BAR, MBAR
\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
\item FEP (free energy peturbation), alchemical, AFE (alchemical free energy)
\end{itemize}
\end{itemize}
%%%%%%%%%%%%%%%%%%%%
%                Software                    %
%%%%%%%%%%%%%%%%%%%%
\section{Available software -- a summary}
\label{sec:software}
\begin{itemize}
\item Commercial:
   \begin{itemize}
    \item FEP+
    \end{itemize}
\item Free or low-cost for academics / commercial for industry:
	\begin{itemize}
	\item CHARMM / DOMDEC / CHARMM-OPENMM
	\item TIES and AMBER FEW? (Peter Coveney)
	\item AMBER / PMEMD
	\end{itemize}
\item Free (libre) open source:
	\begin{itemize}
	\item SIRE
	\item YANK
	\item gromacs
	\item pmx for mutations
	\end{itemize}
\item Setup tools
	\begin{itemize}
	\item FESetup: AMBER, gromacs, Sire
	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
	\end{itemize}
\item Analysis tools:
	\begin{itemize}
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
	\url{https://github.com/michellab/freenrgworkflows}
	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
	\url{https://github.com/choderalab/pymbar}
	\end{itemize}
\end{itemize}

>>>>

\author[1*]{John D. Chodera}
\author[2*]{Antonia S. J. S. Mey}
\author[2*]{Julien Michel}
\author[3*]{David L. Mobley}
\author[4*]{Samarjeet Prasad}
\author[5*]{Julia E. Rice}
\author[6*]{Michael Shirts}
\author[7]{Bryce K. Allen}
\author[1]{Levi N. Naden}
\author[1,9]{Andrea Rizzi}
\affil[1]{Computational and Systems Biology Program, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[2]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[3]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
\affil[4]{I don't know my affiliation}
\affil[5]{I don't know my affiliation}
\affil[6]{University of Colorado Boulder, Boulder, CO, USA}
\affil[7]{Silicon Therapeutics, Boston, MA, USA}
\affil[8]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}

@@

%%%%%%%%%%%%%%%%%%%%
%              Introduction                  %
%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
Alchemical free energy calculations have become a mature technology for computing various properties related to the transfer of chemical species from one environment to another.
The domain of applicability for these calculations now involves such varied applications as the computation of protein-ligand binding free energies~\cite{binding-free-energies}, ligand selectivities~\cite{selectivity}, partition and distribution coefficients between different liquid phases (such as octanol-water partition coefficients~\cite{octanol-water-partition}), loss of affinity due to resistance mutations~\cite{hauser2018predicting,aldeghi2018accurate}, and changes in protein thermostability due to engineered mutations~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}.

The defining characteristic of alchemical free energy calculations is the use of a series of alchemically-modified potential functions $U(x; \lambda)$ in which an alchemical parameter $\lambda$ modulates interactions in a manner that cannot occur in real chemical systems.
One or more simulations are used to collect data from a multitude of alchemical states to compute a free energy difference between a chemical state ($\lambda_0$) and another chemical or alchemical reference state ($\lambda_1$),
\begin{eqnarray}
\Delta f &\equiv& f(\lambda_1) - f(\lambda_0) = - \ln \frac{Z(\lambda_1)}{Z(\lambda_0)} , \label{equation:dimensionless-free-energy-difference}
\end{eqnarray}
where the dimensionless free energy $f(\lambda) \equiv \beta F(\lambda)$ is given in terms of partition functions $Z(\lambda)$,
\begin{eqnarray}
Z(\lambda) &=& \int dx \, e^{-u(x; \lambda)} . \label{equation:partition-function-definition}
\end{eqnarray}
Here, the inverse thermal energy $\beta \equiv (k_B T)^{-1}$ where $k_B$ is the Boltzmann constant and $T$ is the absolute temperature, and the \emph{reduced potential} $u(x; \lambda)$~\cite{shirts2008statistically} is generally given by a trace over thermodynamic parameters with their conjugate dynamical variables,
\begin{eqnarray}
u(x;\lambda) &\equiv& \beta \left[ U(x;\lambda) + p \, V(x) + \sum_{i=1}^N \mu_i \, N_i(x) + \cdots \right] . \label{equation:reduced-potential}
\end{eqnarray}
Here, the collection of thermodynamic and alchemical parameters $\theta \equiv \{\beta, \lambda, p, \mu, \ldots\}$ defines a \emph{thermodynamic state}.
The physical transformation of interest---such as a protein-ligand association process, or transfer free energy among phases---may involve a thermodynamic cycle that requires several free energy differences $\Delta f$ to be computed in order to produce an estimate of the overall desired quantity.

\todo[inline]{JDC: Should we give a few examples, along with illustrative figures?}

@@

In this review, we assume a basic familiarity with the principles of molecular mechanics, molecular dynamics simulations, statistical mechanics, and the biophysics of protein-ligand association.
This document focuses on aspects specific to alchemical free energy calculations, which remain an advanced topic within the molecular simulations area.
Thus, if you are a beginner in the area of molecular simulations, not all background you need will be covered within the scope of this document.
For example, we assume that you are able to set up and conduct a successful equilibrium molecular dynamics simulation of a particular thermodynamic state of a system interest as a prerequisite for this document.

@@

\section{Step 0 -- What can be expected from alchemical simulations?}
\label{sec:step0}
\todo[inline]{Huafeng XU/David Mobely: write section?}
\begin{itemize}
\item What level of accuracy you can expect?
\item What timescales and how many transformations can you address given available computational resources?
\item Can you even hope to tackle the problem you are attempting?
\end{itemize}

\section{Step 1 -- Simulation prerequesites}
\label{sec:step1}
\todo[inline]{ASJSM & JM: write section}
\begin{enumerate}
\item Generate geometry of initial state: Reference biomolecular simulation preparation best practices
\item Relative: Generate geometry of final state (Mey)
\begin{itemize}
\item e.g. ligand ideally should not clash with receptor, etc.; satisfy constraints that might be imposed by protocol such as overlapping atoms, etc.
\item reference biomolecular preparation setup practices for placing ligand into binding site, but elaborate on constraints that must be considered for relative free energy calculations
\end{itemize}
\end{enumerate}

@@

Many protocol issues for alchemical calculations are common, but some are different between absolute and relative calculations, so before treating the common elements we treat the protocol differences.
\todo[inline]{JM: single/dual hybrid toplogies what are they?}
\subsubsection{Relative free energy calculations must select a topology and produce an atom mapping}

@@

\section{Step 3 -- Overview of available analysis techniques}
\label{sec:step3}
\todo[inline, color={red!40}]{JDC, BA: uncertainty estimation section}
\begin{enumerate}
\todo[inline]{JDC: write section}
\item Detecting boundary between equilibrated and production regions (Chodera: \url{http://dx.doi.org/10.1021/acs.jctc.5b00784})
\item Decorrelating samples for analysis
\begin{itemize}
\item Subsample different lambdas based on correlation times
\item Ensure all simulations at least 50x correlation time
\end{itemize}
\todo[inline]{MS: write section}
\item Examining output data for common problems with discussions of what exactly to plot or look at; examples of typical curves for dV/dlambda and free energy versus lambda, for example
\begin{itemize}
\item Make sure ligand doesn’t tumble out of binding site (Mey has observed this)
\item Significant discrepancies between different free energy estimators (TI, BAR, MBAR)
\item Poor replica mixing (for replica-exchange)
\item Correlation time as a function of lambda as it would be expected to be a smooth
\item Dependence on initial conformation
\item Torsional analysis: Is it stuck in specific states? Only very rarely transitions?
\item More “usual suspects”
\end{itemize}
\todo[inline]{MS/JDC: write section}
\item Estimators for free energies
\begin{itemize}
\item MBAR recommended if all energy differences are available
\item BAR just as good for highly optimized lambda values
\item TI should be roughly concordant, but quadrature error hard to quantify
\item Other variants useful in special circumstances (e.g. Z. Tan stochastic version)
\end{itemize}
\todo[inline]{BA: write section}
\item Computing and reporting uncertainties on free energies
correlated bootstrap v. timeseries analysis
\begin{itemize}
\item Quantifying standard error in dG estimate
\item Varience in dG estimate using multiple methods (TI, DEXP, IEXP, BAR, MBAR, GDEL, etc.)
\item Agreement in dG estimate when repeating calculations with different parameters (random seeds, initial configurations, forcefield, etc.)
\item Calculating differences in free energy change as a function of time
\begin{itemize}
	\item Starting from beginning or end of simulation
	\item Significance of differences in midpoint estimate (High middle error: high uncertainty)
\end{itemize}
\item Ensemble method to combine uncertainties into interpretable weighted metric
\begin{itemize}
	\item Simple normalized metric to determine confidence in calculation
	\item Easily interpreted by chemist/biologist when prioritizing new chemistries
\end{itemize}
\end{itemize}
\item Other considerations for many transformations
\todo[inline]{JM: write section}
Cycle closure error

@@

%%%%%%%%%%%%%%%%%%%%
%              Terminology                  %
%%%%%%%%%%%%%%%%%%%%
\section{Terminology and abbreviations}
\label{sec:tem-abbrev}
\begin{itemize}
\item Feature Box covering major technical terms and abbreviations
\item Examples:
\begin{itemize}
\item EXP, BAR, MBAR
\item Double decoupling, single-topology, dual-topology, hybrid-topology, coupled-topology
\item FEP (free energy peturbation), alchemical, AFE (alchemical free energy)
\end{itemize}
\end{itemize}
%%%%%%%%%%%%%%%%%%%%
%                Software                    %
%%%%%%%%%%%%%%%%%%%%
\section{Available software -- a summary}
\todo[inline]{Samar: write section}
\label{sec:software}
\begin{itemize}
\item Commercial:
   \begin{itemize}
    \item FEP+
    \end{itemize}
\item Free or low-cost for academics / commercial for industry:
	\begin{itemize}
	\item CHARMM / DOMDEC / CHARMM-OPENMM
	\item TIES and AMBER FEW? (Peter Coveney)
	\item AMBER / PMEMD
	\end{itemize}
\item Free (libre) open source:
	\begin{itemize}
	\item SIRE
	\item YANK
	\item gromacs
	\item pmx for mutations
	\end{itemize}
\item Setup tools
	\begin{itemize}
	\item FESetup: AMBER, gromacs, Sire
	\item Lomap/Lomap2 : Relative alchemical transformation graph planning
	\end{itemize}
\item Analysis tools:
	\begin{itemize}
	\item Free Energy Workflows: Sire-specific free energy map analysis using weighted path averages
	\url{https://github.com/michellab/freenrgworkflows}
	\item Alchemlyb: Multipackage free energy analysis
	\url{https://github.com/alchemistry/alchemlyb}
	\item pymbar: MBAR implementation, but have to roll your own analysis wrapper
	\url{https://github.com/choderalab/pymbar}
	\end{itemize}
\end{itemize}
